---
title: "An investigation into the variables that impact mental health"
author: "Kylen Brown (3540337), Akira Govender (3540386) & Shaundre Hofstander (3540248)"
date: "28 June 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3. Suicide risk 

# 3.1. The effect of depressive symptoms intensity on suicide risk 

## Using a Wilcox test for the non-normal data instead of a t-test

```{r}
wilcox.test(CESD ~ Suicide_risk, data = health)
```

# 3.2. The effect of social support on suicide risk 

## Using a Wilcox test for the non-normal data instead of a t-test

```{r}
wilcox.test(Social_support ~ Suicide_risk, data = health)
```

# 4. Overall analysis

## Creating a correlation plot

### Loading data 

```{r}
health_2 <- read.csv2("health_edit.csv")
```

```{r}
summary(health_2)
```

### Getting N and P values

```{r}
h2 <- rcorr(as.matrix(health_2, use = "complete obs"))
h2
```

```{r}
dat1 <- health_2[sapply(health_2,function(x) length(unique(x)))>1]
cor(dat1, use = "complete.obs")
```

### Creating a function for the cor matrix

```{r}
flattenCorrMatrix <- function(cormat, pmat) 
  {ut <- upper.tri(cormat)
  data.frame(row = rownames(cormat)[row(cormat)[ut]],
             column = rownames(cormat)[col(cormat)[ut]],
             cor  =(cormat)[ut],
             p = pmat[ut])}
```

### Creating correlation matrix

```{r}
final <- cor(health_2, use = "complete.obs")
final
```

### Creating the visualisation

```{r}
corrplot(final, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```

