---
title: "An investigation into the variables that impact mental health"
author: "Kylen Brown (3540337), Akira Govender (3540386) & Shaundre Hofstander (3540248)"
date: "28 June 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Setup 

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggpubr)
library(readr)
library(rmarkdown)
library(knitr)
library(e1071)
library(ggplot2)
library(magrittr)
library(Hmisc)
library(corrplot)
library(RColorBrewer)
library(ggthemes)
library(DT)
library(xtable)
library(devtools)
```
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the data and assigning to an object

```{r}
health <- read.csv2("health.csv")
```

# 1. Depressive symptoms (CESD)

## 1.1 Hypothesis 1: The effect of different substances on depressive symptoms 

HO: Substance has no effect on CESD 
H1: Substance has an effect on CESD

## Creating a graph representing the effect of each substance on the intenisty of depressive symptoms experienced

### Group and mutate

```{r}
depression_data <- health %>%
  group_by(Substance) %>%
  summarise(mn_cesd = mean(CESD, na.rm = TRUE),
            sd_cesd = sd(CESD, na.rm = TRUE))
```
This created an average for the 3 substances that were abused

### Creating a table

```{r}
kable(depression_data, caption = "Table 1. The effect of each substance on depressive symptoms")
```

### Creating the graph
```{r}
sub_depression_graph <- ggplot(depression_data, aes(x = Substance, y = mn_cesd, 
                                              fill = Substance)) +
  geom_col(aes(fill = Substance), position = "dodge",width = 0.5) +
  labs(x = "Substance",y = "Average measure of depression",
       title = "The effect of each substance on the intensity of depressive symptoms", 
       caption = "Figure 1: The average measure of depressive symptoms based on the three substances that were abused") +
  theme(plot.title = element_text(hjust = 0.5), 
        plot.caption = element_text(hjust = 0.5)) +
  geom_errorbar(aes(ymin = mn_cesd - sd_cesd,
                    ymax = mn_cesd + sd_cesd),
                position = "dodge",width = 0.5)
sub_depression_graph
```

## Comaparing the difference in mean and median against skewness and calculating the kurtosis of the tails of a distribution (depressive symptoms based on substances abused) 

```{r}
health %>%
  group_by(Substance) %>% 
  summarise(mean_cesd = mean(CESD),
            median_cesd = median(CESD),
            skew_cesd = skewness(CESD),
            kurt_cesd = kurtosis(CESD))
```

## Conducting an ANOVA test 

```{r}
CESD.aov <- aov(CESD ~ Substance, data = health)
summary(CESD.aov)
```

## Creating the graphical representation

```{r}
ggplot(data = CESD.aov, aes(x = Substance, y = CESD)) +
  geom_boxplot(aes(fill = Substance), notch = TRUE) +
  labs(title = "The effect of each substance on the intensity of depressive symptoms", 
       caption = "Figure 2: An ANOVA analysis of the depressive symptoms based on the three substances that were abused") +
  theme(plot.title = element_text(hjust = 0.5), 
        plot.caption = element_text(hjust = 0.5))
```

## Running a tukey test
 
```{r}
health_Tukey <- as.data.frame(TukeyHSD(aov(CESD ~ Substance, data = health))$Substance)
health_Tukey$pairs <- as.factor(row.names(health_Tukey))
summary(health_Tukey)
```

## 1.2. Hypothesis 2: The effect of gender on depressive symptoms 

HO: Gender has no effect on CESD 
H1: Gender has an effect on CESD

## Comaparing the difference in mean and median against skewness and calculating the kurtosis of the tails of a distribution (depressive symptoms based on sex) 

```{r}
health %>%
  group_by(Sex) %>% 
  summarise(mean_cesd2 = mean(CESD),
            median_cesd2 = median(CESD),
            skew_cesd2 = skewness(CESD),
            kurt_cesd2 = kurtosis(CESD))
```

## Conducting an ANOVA test 

```{r}
CESD2.aov <- aov(CESD ~ Sex, data = health)
summary(CESD2.aov)
```

# 1.3. The influence of age on the intensity of depressive symptoms   

## Running the correlation test between depressive symptoms and age 

```{r}
cor.test(health$CESD, health$Age, method = "pearson",
         conf.level = 0.95)
```

# 2. Mental component score

## 2.1 Hypothesis 3: The effect of different substances on mental component score 

HO: Substance has no effect on Mental component score 
H1: Substance has an effect on Mental component score

## Creating a graph representing the effect of each substance on mental component score

### Group and mutate

```{r}
mental_data <- health %>%
  mutate(Mental_score = as.numeric(Mental_score)) %>% 
  group_by(Substance) %>%
  summarise(mn_mental = mean(Mental_score, na.rm = TRUE),
            sd_mental = sd(Mental_score, na.rm = TRUE))
```
This created an average for the 3 substances that were abused

### Creating a table

```{r}
kable(mental_data, caption = "Table 2. The effect of each substance on mental score")
```

### Creating the graph

```{r}
mental_score_graph <- ggplot(mental_data, aes(x = Substance, y = mn_mental, 
                                              fill = Substance)) +
  geom_col(aes(fill = Substance), position = "dodge",width = 0.5) +
  labs(x = "Substance",y = "Average mental component score",
       title = "The effect of each substance on the mental component score", 
       caption = "Figure 3: The average mental compnent score based on the three substances that were abused") +
  theme(plot.title = element_text(hjust = 0.5), 
        plot.caption = element_text(hjust = 0.5)) +
  geom_errorbar(aes(ymin = mn_mental - sd_mental,
                    ymax = mn_mental + sd_mental),
                position = "dodge",width = 0.5)
mental_score_graph
```

## 2.2. Hypothesis 4: The influence of gender on mental component score 

HO: Gender has no effect on Mental component score 
H1: Gender has an effect on Mental component score

## Creating a graph representing mental component score for both males and females

### Group and mutate

```{r}
gender_data <- health %>%
  mutate(Mental_score = as.numeric(Mental_score)) %>% 
  group_by(Sex) %>%
  summarise(mn_mental = mean(Mental_score, na.rm = TRUE),
            sd_mental = sd(Mental_score, na.rm = TRUE))
```
This compared the average mental score in terms of gender

### Creating the graph
```{r}
gender_graph <- ggplot(gender_data, aes(x = Sex, y = mn_mental, 
                                              fill = Sex)) +
  geom_col(aes(fill = Sex), position = "dodge",width = 0.5) +
  labs(x = "Gender",y = "Average mental component score",
       title = "Average mental component score of males and females", 
       caption = "Figure 4: The average mental component score based upon gender" ) +
  theme(plot.title = element_text(hjust = 0.5), 
        plot.caption = element_text(hjust = 0.5)) +
  geom_errorbar(aes(ymin = mn_mental - sd_mental,
                    ymax = mn_mental + sd_mental),
                position = "dodge",width = 0.5)
gender_graph
```

##  Measures of Variance

### A summary of the statistical properties 

```{r}
ms_summary2 <- health %>% 
  group_by(Sex) %>% 
  summarise(mean_ms2 = mean(Mental_score),
            median_ms2 = median(Mental_score),
            var_ms2 = var(Mental_score),
            sd_ms2 = sd(Mental_score),
            min_ms2 = min(Mental_score),
            quart1_ms2 = quantile(Mental_score, 0.25),
            quart2_ms2 = quantile(Mental_score, 0.5),
            quart3_ms2 = quantile(Mental_score, 0.75))
ms_summary2
```

## Conducting an ANOVA test 

```{r}
mentals.aov <- aov(Mental_score ~ Sex, data = health)
summary(mentals.aov)
```

# 2.3. The effect of social support on mental component score 

## Running a correlation test between Mental health score and social support

```{r}
cor.test(health$Mental_score, health$Social_support, method = "pearson",
         conf.level = 0.95)
```
# 2.4 The effect of age on mental component score

## A correlation analysis 

### Group and mutate

```{r}
age_data <- health %>%
  mutate(Mental_score = as.numeric(Mental_score)) %>% 
  group_by(Age) %>%
  summarise(mn_mental = mean(Mental_score, na.rm = TRUE),
            sd_mental = sd(Mental_score, na.rm = TRUE))
```
This compared the average mental score in terms of age

### Creating the graph

```{r}
correlation <- ggplot(data = age_data, aes(x = Age , y = mn_mental)) +
  geom_point(pch = 16, cex = 1.5, colour = "Navy") +
  geom_smooth(method = lm, colour = "Red", se = FALSE) +
  labs(x = "Age", y = "Mental component score", 
       caption = "Figure 5: Correlation between age and the mental component score") +
  ggtitle("Average mental component score based on age") +
  theme(plot.title = element_text(hjust = 0.5), 
        plot.caption = element_text(hjust = 0.5))
correlation
```





